Demystifying the Computer Science PhD Admission in the US\\{\large A Guide for International Students

### 2023-12-23 demystify.tex
updates

Additionally, \textbf{work experiences at renowned research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you should explicit say something about them in your statement.

@@

Finally, \textbf{participating internationally recognized competitions} can also demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or  winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.

@@

\section{Grades/GREs}\label{sec:grades}

Having good grades is good, but unless your school is well-known, having top grades or rankings
usually will not help because we cannot evaluate them.

This can be an issue for students in many top international universities where the competition is so high that very good students can still have low rankings from these schools (and be overlooked by the admission committee).
So what to do with this? as mentioned in \S\ref{sec:your-school} you can put a note about this in your statement and ask your LoR writers to talk about it.

Note that while having good grades at unknown school might not help,
having bad grades will be \red{red flag} (unless your LORs or
statement give a proper explanation). This is especially true if you
have bad grades in relevant courses (e.g., Math and CS).

@@

\section{CV/Resume}
The CV/Resume should provide a summary of the applicant's achievements.  It should allow the reviewers to quickly scan to identify major achievements (e.g., Publications, Programming Competition Awards, Teaching Experience).

@@

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider are the reputation of schools and professors (\S\ref{sec:schoolsandprofs}), and funding availability (\S\ref{sec:funding}). You will have to make your decision by around \emph{April 15}.

\paragraph{Open House} Most schools will have \textbf{Open House} events, which are a great resource to learn about the school, department, faculty, research, living, etc. Even if you can't come in person, you should attend virtually and meet with individual faculty.
During the Open House, you get a chance to talk to individual faculty and current students.  Take notes of faculty who make you excited, count those that are taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisers, the dept, the area, funding situation etc.  Ask about anything you want to determine that they deserve \emph{you}.

@@

\paragraph{If you do not get admitted} If you do not get admitted to any schools or don't want to go to the ones that admit you, try again next time.  Graduate admission can involve randomness and noise.  Don't bother asking for feedback, you will not likely get any.  In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for better LoR writers, etc.

>>>>

Additionally, \textbf{work experiences at renowned research laboratories}, such as Microsoft Research, can significantly strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you should explicit say something about them in your statement.

@@

Finally, \textbf{participating internationally recognized competitions} can also demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or  winning ACM programming contests if you want to ``build'' stuff, e.g., software analysis.

@@

\section{Grades and GREs}\label{sec:grades}

Having good grades is important, but unless your school is well-known, having top grades or rankings
usually will not help because we cannot evaluate them.

This can be an issue for students in many top international universities where the competition is so high that very good students students can still have low rankings from these schools (and be overlooked by the admission committee).
So what to do with this? as mentioned in \S\ref{sec:your-school} you can put a note about this in your statement and ask your LoR writers to talk about it.

Note that while having good grades at unknown school might not help,
having very bad grades will be \red{red flag} (unless your LORs or
statement give a proper explanation). This is especially true if you
have bad grades in relevant courses (e.g., Math and CS).

@@

\section{CV/Resume}
The CV/Resume should provide a summary of the applicant's achievements.  It should allow the reviewers to quickly scan to identify standout achievements (e.g., Publications, Programming Competition Awards, Teaching Experience).

@@

By around March you should hear back from most PhD programs you applied to. If you haven't heard back, reach out via email and ask about the status of your application.
If you receive offers, congratulations!  Now you're at a different game because the schools that have admitted you will try to get you to accept them!  Important factors to consider are the reputation of schools and professors (\S\ref{sec:schoolsandprofs}), and of course, funding availability (\S\ref{sec:funding}). You will have to make your decision by around \emph{April 15}.

\paragraph{Open House} Most schools will have \textbf{Open House} events, which are a great resource to learn about the school, department, faculty, research, living, etc. Even if you can't come in person, you still can attend virtually and meet with individual faculty.
During the Open House, you get a chance to talk to individual faculty and current students.  Take notes of faculty who make you excited, count those that are taking in new students (if they meet you, likely they are considering new students!).  Talk to students about their advisers, the dept, the area, funding situation etc.  Ask about anything you want to determine that they deserve \emph{you}.

@@

\paragraph{If you do not get admitted} If you do not get admitted to any schools (or don't want to go to the ones that admit you), try again next time.  Graduate admission can involve randomness and noise.  Don't bother asking for feedback, you will not likely get any.  In the meantime, you can work on improving your profile, e.g., get more research experiences, publish more papers, improve your connections for better LoR writers, etc.
